The four postulates of quantum mechanics are three

### 2020-12-05 Articles/2019-QuantumComposition/Submissions/PhysLetters/tensor.tex

Let us first provide a conceptual overview of our approach. (XXX) We start from the natural definition of a composite system as the set
of two (or more) quantum systems. The composite system is therefore
made of system $A$ {\em and} (joined with) system $B$ and {\em nothing
  else}. The first key insight is that the first two postulates of
quantum theory (introduced below) already assume that the preparation
of one system is independent from the preparation of another
(statistical independence). In fact, we cannot even talk about a
system in the first place if we cannot characterize it independently.
The second key insight is that, since all measurement outcomes and all
states of the subsystems plus the composite live in the same
probability space, we will automatically have a map $M$ that for a
pair of subsystems' states gives us the composite state that
corresponds to the statistically independent case. These insights are
enough to characterize mathematically the state space of the
composite: the linearity given by the Hilbert space, together with the
fact that the composite system is fully described by the observables
of $A$ and $B$, allows us (XXX) to extend the construction to the composite
states that do not satisfy statistical independence (the entangled
states). So the work consists of two interrelated efforts: a physical
argument that starts from the first two postulates and leads to the
necessary existence of the composition map $M$ and its properties
together with a formal argument that shows how $M$ leads to the tensor
product.

@@

This map $M$ acts on the state spaces of the subsystems. Each pure
state is identified by a ray $\pj{\psi}$, a subspace of the system's
Hilbert space comprising (XXX) all vectors $\psi$ differing by their
(nonzero) modulus and phase: a one-dimensional complex subspace (a complex plane). In
the same way, constraining the observable $X$ to a particular outcome
value $x_0$ means identifying the subspace comprising (XXX) all
non-normalized eigenvectors $|x_0\>$ of arbitrary phase such that
$X | x_0 \> = x_0 |x_0\>$. As we need to map states and observables
between subsystems and composite system, the map $M$ derived from the
probability space is really a map between subspaces, namely the
projective spaces. We then use projective geometry to show that the
map $M$ on the projective spaces corresponds to a map $m$ on the
vector spaces which is the tensor product.

This is done by showing that preparation independence and statistical
independence of the systems imply three conditions on the map $m$:
(H1)~totality: the map is defined on all states of the subsystems;
(H2)~bilinearity: the map is bilinear thanks to the fundamental
theorem of projective geometry; (H3)~span surjectivity: the span of
the image of the map coincides with the full composite Hilbert space.  We
then prove that, if these three conditions H1, H2 and H3 hold, then
the map $m$ is the tensor product, namely the Hilbert space of the
composite system is the (XXX) tensor product of the components' Hilbert spaces (XXX). The tensor
product ``postulate'' hence loses its status of a postulate. (XXX) An
overview of all these logical implications is given in
Fig.~\ref{f:fig}. The rest of the paper contains the sketch of this
argument, including all the physical arguments outlined above. The
supplementary material contains the mathematical details.

@@

Step 2: use the universal property. The tensor product is uniquely
characterized, up to isomorphism, by a universal property regarding
bilinear maps: given two vector spaces $\cal A$ and $\cal B$, the
tensor product ${\cal A}\otimes{\cal B}$ and the associated bilinear
map $T : \cal A \times \cal B \to {\cal A}\otimes{\cal B}$ have the
property than any bilinear map $m:{\cal A}\times{\cal B}\to{\cal C}$
factors through $T$ uniquely.  This means that there exists a {\em
  unique} $I$, dependent on $m$, such that $I \circ T=m$.
In other words, the following diagram commutes:
\begin{center}
		\begin{tikzcd}\mathcal{A}\times\mathcal{B} \arrow[rd, "m"]\arrow[r, "T"] & \mathcal{A}\otimes\mathcal{B}\arrow[d, "I"] \\
			& \mathcal{C}
		\end{tikzcd}
\end{center}
Since $m : \mathcal{A} \times \mathcal{B} \to \mathcal{C}$ is a
bilinear operator (property H2), thanks to the universal property of
the tensor product we can find a unique linear operator
$I : \mathcal{A} \otimes \mathcal{B} \to \mathcal{C}$ such that
$m(a, b) = I(a \otimes b)$. The set $\{ I(a_i\otimes b_j)$
with $|a_i\>$ and $|b_j\>$ orthonormal bases for $\cal A$ and
${\cal B}\}$ forms a basis for $\cal C$, since
$I(a_i\otimes b_j)=m(a_i,b_j)$ and we have shown above that the
latter is a basis.  Thus,
\begin{align} 
  &\<I(a_i\otimes b_j)|I(a_k\otimes b_\ell)
    \>_{\mathcal{C}}=\<m(a_i,
    b_j)|m(a_k,b_\ell)\>_\mathcal{C} \nonumber\\& =
                                                  \delta_{ik}\delta_{j\ell}
                                                  = \<a_i\otimes
                                                  b_j| a_k \otimes b_\ell\>_{\otimes},
	\labell{ecco}\; 
\end{align}
where we used the orthonormality of the bases and the fact that
$|a_i\otimes b_j\>$ is a basis of the tensor product space
${\cal A}\otimes{\cal B}$. Since the function $I$ is a linear
function that maps an orthonormal basis of ${\cal A}\otimes{\cal B}$
to an orthonormal basis of $\cal C$, $I$ is a an isomorphism (a
bijection that preserves the mathematical structure) between
${\cal A}\otimes{\cal B}$ and $\cal C$. As
${\cal C}\cong{\cal A}\otimes{\cal B}$ are isomorphic as Hilbert
spaces, they are mathematically equivalent: $c \in {\cal C}$ and $I^{-1}(c)$ (XXX) represent the same physical object. In this sense, we can loosely say that $I$ is the identity, as it connects spaces that are physically 
equivalent. So we can directly use the
tensor product to represent the composite state space. This means that
the map $m : \cal A \times \cal B \to \cal C$ is equivalent to the map
$\otimes : \cal A \times \cal B \to \cal A \otimes \cal B$ in the
sense that $m \circ I^{-1} = \otimes$.$\square$

@@

\vskip 1\baselineskip A few comments on the proof: it is based on the
universal property of the tensor product, which uniquely characterizes
it. In step 1 we show that the bilinear map $m$ maps subsystems' bases
into the composite system basis. We also know that there exists a
tensor product map $ {T}=\otimes$ that can compose the vectors in
$\cal A$ and $\cal B$.  In step 2 we use the universal property: since
$m$ is a bilinear map, we are assured that there exists a unique
$I$ such that $I \circ T=m$. Since we show that $I$ is
an isomorphism, then $I$ bijectively maps vectors in $\cal C$
onto vectors in the tensor product space. Namely $m={T}=\otimes$. (XXX)

>>>>

Let us first provide a conceptual overview of our approach. We start from the natural definition of a composite system as the set
of two (or more) quantum systems. The composite system is therefore
made of system $A$ {\em and} (joined with) system $B$ and {\em nothing
  else}. The first key insight is that the first two postulates of
quantum theory (introduced below) already assume that the preparation
of one system is independent from the preparation of another
(statistical independence). In fact, we cannot even talk about a
system in the first place if we cannot characterize it independently.
The second key insight is that, since all measurement outcomes and all
states of the subsystems plus the composite live in the same
probability space, we will automatically have a map $M$ that for a
pair of subsystems' states gives us the composite state that
corresponds to the statistically independent case. These insights are
enough to characterize mathematically the state space of the
composite: the linearity given by the Hilbert space, together with the
fact that the composite system is fully described by the observables
of $A$ and $B$, allows us to extend the construction to the composite
states that do not satisfy statistical independence (the entangled
states). So the work consists of two interrelated efforts: a physical
argument that starts from the first two postulates and leads to the
necessary existence of the composition map $M$ and its properties
together with a formal argument that shows how $M$ leads to the tensor
product.

@@

This map $M$ acts on the state spaces of the subsystems. Each pure
state is identified by a ray $\pj{\psi}$, a subspace of the system's
Hilbert space comprising all vectors $\psi$ differing by their
(nonzero) modulus and phase: a one-dimensional complex subspace (a complex plane). In
the same way, constraining the observable $X$ to a particular outcome
value $x_0$ means identifying the subspace comprising all
non-normalized eigenvectors $|x_0\>$ of arbitrary phase such that
$X | x_0 \> = x_0 |x_0\>$. As we need to map states and observables
between subsystems and composite system, the map $M$ derived from the
probability space is really a map between subspaces, namely the
projective spaces. We then use projective geometry to show that the
map $M$ on the projective spaces corresponds to a map $m$ on the
vector spaces which is the tensor product.

This is done by showing that preparation independence and statistical
independence of the systems imply three conditions on the map $m$:
(H1)~totality: the map is defined on all states of the subsystems;
(H2)~bilinearity: the map is bilinear thanks to the fundamental
theorem of projective geometry; (H3)~span surjectivity: the span of
the image of the map coincides with the full composite Hilbert space.  We
then prove that, if these three conditions H1, H2 and H3 hold, then
the map $m$ is the tensor product, namely the Hilbert space of the
composite system is the tensor product of the components' Hilbert spaces. The tensor
product ``postulate'' hence loses its status of a postulate. An
overview of all these logical implications is given in
Fig.~\ref{f:fig}. The rest of the paper contains the sketch of this
argument, including all the physical arguments outlined above. The
supplementary material contains the mathematical details.

@@

Step 2: use the universal property. The tensor product is uniquely
characterized, up to isomorphism, by a universal property regarding
bilinear maps: given two vector spaces $\cal A$ and $\cal B$, the
tensor product ${\cal A}\otimes{\cal B}$ and the associated bilinear
map $T : \cal A \times \cal B \to {\cal A}\otimes{\cal B}$ have the
property than any bilinear map $m:{\cal A}\times{\cal B}\to{\cal C}$
factors through $T$ uniquely.  This means that there exists a {\em
  unique} $I$, dependent on $m$, such that $I \circ T=m$.
In other words, the following diagram commutes:
\begin{center}
		\begin{tikzcd}\mathcal{A}\times\mathcal{B} \arrow[rd, "m"]\arrow[r, "T"] & \mathcal{A}\otimes\mathcal{B}\arrow[d, "I"] \\
			& \mathcal{C}
		\end{tikzcd}
\end{center}
Since $m : \mathcal{A} \times \mathcal{B} \to \mathcal{C}$ is a
bilinear operator (property H2), thanks to the universal property of
the tensor product we can find a unique linear operator
$I : \mathcal{A} \otimes \mathcal{B} \to \mathcal{C}$ such that
$m(a, b) = I(a \otimes b)$. The set $\{ I(a_i\otimes b_j)$
with $|a_i\>$ and $|b_j\>$ orthonormal bases for $\cal A$ and
${\cal B}\}$ forms a basis for $\cal C$, since
$I(a_i\otimes b_j)=m(a_i,b_j)$ and we have shown above that the
latter is a basis.  Thus,
\begin{align} 
  &\<I(a_i\otimes b_j)|I(a_k\otimes b_\ell)
    \>_{\mathcal{C}}=\<m(a_i,
    b_j)|m(a_k,b_\ell)\>_\mathcal{C} \nonumber\\& =
                                                  \delta_{ik}\delta_{j\ell}
                                                  = \<a_i\otimes
                                                  b_j| a_k \otimes b_\ell\>_{\otimes},
	\labell{ecco}\; 
\end{align}
where we used the orthonormality of the bases and the fact that
$|a_i\otimes b_j\>$ is a basis of the tensor product space
${\cal A}\otimes{\cal B}$. Since the function $I$ is a linear
function that maps an orthonormal basis of ${\cal A}\otimes{\cal B}$
to an orthonormal basis of $\cal C$, $I$ is a an isomorphism (a
bijection that preserves the mathematical structure) between
${\cal A}\otimes{\cal B}$ and $\cal C$. As
${\cal C}\cong{\cal A}\otimes{\cal B}$ are isomorphic as Hilbert
spaces, they are mathematically equivalent: $c \in {\cal C}$ and $I^{-1}(c)$ represent the same physical object. In this sense, we can loosely say that $I$ is the identity, as it connects spaces that are physically 
equivalent. So we can directly use the
tensor product to represent the composite state space. This means that
the map $m : \cal A \times \cal B \to \cal C$ is equivalent to the map
$\otimes : \cal A \times \cal B \to \cal A \otimes \cal B$ in the
sense that $m \circ I^{-1} = \otimes$.$\square$

@@

\vskip 1\baselineskip A few comments on the proof: it is based on the
universal property of the tensor product, which uniquely characterizes
it. In step 1 we show that the bilinear map $m$ maps subsystems' bases
into the composite system basis. We also know that there exists a
tensor product map $ {T}=\otimes$ that can compose the vectors in
$\cal A$ and $\cal B$.  In step 2 we use the universal property: since
$m$ is a bilinear map, we are assured that there exists a unique
$I$ such that $I \circ T=m$. Since we show that $I$ is
an isomorphism, then $I$ bijectively maps vectors in $\cal C$
onto vectors in the tensor product space. Namely $m={T}=\otimes$.
